*Salmonella* Environmental Persistence Informs Management Relevant to Avian and Public Health

### 2023-05-05 3. Analysis Script/StatisticalAnalysis.qmd
More updates

# **Analysis script**
This script loads the cleaned data, does a simple analysis, and saves the results to the "4. Results" folder

## **Loading Packages**
```{r}
#Load needed packages
library(tidyverse)
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(dotwhisker)
library(here) #for data loading/saving
library(tidymodels)
library(tidyr)
library(gtsummary)
library(dplyr)
library(vip)
library(performance)
library(jtools)
library(broom)
library(broom.mixed)
```

## **1. Loading: Picnic Table & Feeder Data**
```{r}
#Picnic Table Data
PT_Data<- readRDS(here::here("2. Clean Data","PTprocessed.rds"))

@@

glimpse (FD_Data)
```
# 2. **Data Fitting/Statistical Analysis: Picnic Table**
Let's Begin with some Linear Model Fitting; I will try this multiple ways, with several predictors.
```{r}
lmog<-linear_reg() %>% 
  set_engine("lm")

model_lm<- linear_reg()

lmfit_PT<- model_lm %>% #Training; outcome: Salmonella Pos and Main Predictor Table
  fit(Salmonella_Positive~Date, PT_Data)

lmfit_PT

glimpse(lmfit_PT)

#Way 1
glance(lmfit_PT)

#Way 2
tidy(lmfit_PT)

@@

tidy(lmfit_PT) %>%
    dwplot(dot_args = list(size = 2, color = "coral1"),
         whisker_args = list(color = "coral1"),
         vline = geom_vline(xintercept = 0, color = "darkred", linetype = 4))
```

```{r}
#Fitting other specific indicators
lmog_1<-linear_reg() %>% 
  set_engine("lm")

model_lm1<- linear_reg()

lmfit_PT1<- model_lm1 %>% #Training; outcome: Salmonella Pos
  fit(Salmonella_Positive~Sample_Type, PT_Data)

lmfit_PT1

tidy(lmfit_PT1)

glimpse(lmfit_PT1)

#Way 1
glance(lmfit_PT1)

#Way 2
tidy(lmfit_PT1)

@@

@@

lmall_PT<- lmog_PT %>% #Training 
  fit(Salmonella_Positive~., data=PT_Data)

@@

tidy(lmall_PT)

#Do not want all predictors...

#Checking model performance: Picnic Table
check_model(lmall_PT$fit)
```

@@

#A different way Part Deux 
# I will do the same as above, however, I am now adding persistence pile
lmfit2 <- lm(Salmonella_Positive~Original_Persistence_Pile+Table, PT_Data) 
plot(lmfit2)

@@

Second model fit
```{r}
#Doing the same for FEEDER DATA
lmog_FD<-linear_reg() %>% 
  set_engine("lm")

lmfit_FD<- lmog_FD %>% #Training; outcome: Salmonella Pos and Main Predictor Table
  fit(Absence_0_Presence_1~Feeder_number, FD_Data)

lmfit_FD

@@

#Way 1
glance(lmfit_FD)

#Way 2
tidy(lmfit_FD)

@@

#Checking model performance: FEEDER
check_model(lmfit_FD$fit)

#More fitting with some other indicators

lmog_FD2<-linear_reg() %>% 
  set_engine("lm")

lmfit_FD2<- lmog_FD2 %>% #Training; outcome: Salmonella Pos and Main Predictor Table
  fit(Absence_0_Presence_1~Feeder_number+Feeder_Type, FD_Data)

lmfit_FD2

@@

#Way 1
glance(lmfit_FD2)

#Way 2
tidy(lmfit_FD2)

@@

#Trying ALL predictors of Feeder dataset 
lmog_FD<- linear_reg() %>%
  set_engine("lm")

@@

tidy(lmall_FD)

plot_summs(lmall_FD, robust=TRUE) #KP NOTE: I need to clean this up so it is a bit more readable, must fullscreen it
```
# **Generalized Linear Model: Picnic Table** 
```{r}
lmog<-linear_reg() %>% 
  set_engine("glm")

@@

#Way 1
glance(lmfit_PT_glm)

#Way 2
tidy(lmfit_PT_glm)

@@

@@

lmfit_PTglm1<- lmog_glm1 %>% #Training; outcome: Salmonella Pos and Main Predictor Table
  fit(Salmonella_Positive~Original_Persistence_Pile+Table, PT_Data)

@@

glimpse(lmfit_PTglm1)

#Way 1
glance(lmfit_PTglm1)

#Way 2
tidy(lmfit_PTglm1)

@@

@@

lmfit_PTglm2<- lmog_glm2 %>% #Training; outcome: Salmonella Pos and Main Predictor Table
  fit(Salmonella_Positive~Quadrant+Table, PT_Data) 
```

@@

#Way 1
glance(lmfit_PTglm2)

#Way 2
tidy(lmfit_PTglm2)

@@

@@

# **Generalized Linear Model: Picnic Table (Persistence & Prevalence)** 
```{r}
PT_GLM_PER <- glm(Salmonella_Positive ~ Date, data = PT_Data, family = binomial)

@@

PT_GLM_PREV <- glm(Salmonella_Positive ~ Sample_Type + as.factor(Date) + Table, data= PT_Data, family = binomial)

PT_GLM_PREV
```

# **Generalized Linear Model: Feeder Data**
```{r}
lmog_FD<-linear_reg() %>% 
  set_engine("glm")

@@

#Way 1
glance(lmfit_FD_glm)

@@

@@

#Way 1
glance(lmfit_FD_GLM)

#Way 2
tidy(lmfit_FD_GLM)

@@

@@

tidy(lmall_FDGLM_all1)
```

# **Generalized Linear Model: Feeder Data (Persistence & Prevalence)**
```{r}
FD_GLM_PER <- glm(Absence_0_Presence_1 ~ Collection_Date, data = FD_Data, family = binomial)

@@

FD_GLM_PREV <- glm(Absence_0_Presence_1 ~ Feeder_Type + as.factor(Collection_Date), data= FD_Data, family = binomial)

FD_GLM_PREV
```
# **Time to event**
```{r}
library(finalfit)
library(forcats)
library(survival)

@@

#**Cox regression: Picnic Table Data**
```{r}
PT_Data$Date<- as.numeric(PT_Data$Date)

@@

coxph(Surv(Date, Salmonella_Positive) ~ Sample_Type, PT_Data) %>%
  tbl_regression(exp=TRUE)
```
#**Cox regression: Feeder Data**
```{r}
library(lubridate)

>>>>

# **Analysis script**

This script loads the cleaned data, does a series of analysis and saves the results to the "4. Results" folder.

## **Loading Packages**

```{r}
#Load needed packages
library(tidyverse)
library(ggplot2) #For plotting/data vis.
library(broom) #For cleaning up output from lm()
library(dotwhisker) #Visualization for regression outputs
library(here) #Setting paths/for data loading/saving
library(tidymodels) #Great package for models and other resources used for fitting
library(tidyr) #Helps with data wrangling
library(gtsummary) #To create tables 
library(dplyr)
library(vip)
library(performance) #Model evaluation
library(jtools)
library(broom)
library(broom.mixed) #Converts fitted to tiddy dfs
```

After reviewing both the Picnic Table and Feeder Trial datasets, I will run several analysis including linear regression, generalized linear model, time to event, and cox regression. Given that both datasets are exploring presence/absence of *Salmonella*, I believe the above models will provide the most flexibility with the continuous and categorical predictors contained in the data.

# **1. Loading: Picnic Table & Feeder Data**

Now that the packages we need are loaded, let's read in and view our cleaned data

```{r}
#Picnic Table Data
PT_Data<- readRDS(here::here("2. Clean Data","PTprocessed.rds"))

@@

glimpse (FD_Data)
```

# 2. **Data Fitting/Statistical Analysis: Linear Regression \[Picnic Table Trial\]**

Let's Begin with some Linear Model Fitting; I will try this multiple ways, with several predictors.

To begin fitting, we will first need to define the linear regression.

```{r}
lmog<-linear_reg() %>% 
  set_engine("lm") #Setting

model_lm<- linear_reg()
```

Now that we have defined our linear regression (LR), we will have to train the model to our given data.

```{r}
lmfit_PT<- model_lm %>% #Training outcome: Salmonella Positive and Main Predictor Date
  fit(Salmonella_Positive~Date, PT_Data)

lmfit_PT #With this function we can view what we just did

tidy(lmfit_PT) #We can use the tidy() function to produce the output/ summary statistics

@@

#Plotting the above 
tidy(lmfit_PT) %>%
    dwplot(dot_args = list(size = 2, color = "coral1"),
         whisker_args = list(color = "coral1"),
         vline = geom_vline(xintercept = 0, color = "darkred", linetype = 4))
```

Again, we will need to define the linear regression for fitting of other specific indicators.

```{r}
#Fitting other specific indicators
lmog_1<-linear_reg() %>% 
  set_engine("lm") #Setting

model_lm1<- linear_reg()
```

We have defined our LR, now we will train the model to our given data and new predictor.

```{r}
lmfit_PT1<- model_lm1 %>% #Training; outcome: Salmonella Pos
  fit(Salmonella_Positive~Sample_Type, PT_Data)

lmfit_PT1 #Let's view what we just did

tidy(lmfit_PT1) #We can use the tidy() function to produce the output/ summary statistics

@@

We will follow the same workflow as above, however, now we will utilize all predictors of the Picnic Table Trial Data.

@@

model_lmall<- linear_reg()

lmall_PT<- model_lmall %>% #Training 
  fit(Salmonella_Positive~., data=PT_Data)

@@

tidy(lmall_PT)
```

I do not believe all predictors will be beneficial in this analysis. However, we will continue to check the model performance and include it in the supplemental section.

```{r}
#Checking model performance: Picnic Table
check_model(lmall_PT$fit)
```

The code below \[lines132-154\] will likely be included in the supplemental section as it should produce similar values as the above code. I wanted to include this as an alternate way of writing the code and most importantly a resource to refer back to in the future.

@@

#A different way 
# I will do the same as above, however, I am now adding persistence pile
lmfit2 <- lm(Salmonella_Positive~Original_Persistence_Pile+Table, PT_Data) 
plot(lmfit2)

@@

# 2. **Data Fitting/Statistical Analysis: Linear Regression \[Feeder Trial\]**

This section is a continuation of what was produced above, however, we will now be utilizing the Feeder Trial Data

```{r}
#Doing the same for FEEDER DATA
lmog_FD<-linear_reg() %>% 
  set_engine("lm")

model_lmFD<- linear_reg()

lmfit_FD<- model_lmFD %>% #Training; outcome: Salmonella Pos and Main Predictor Table
  fit(Absence_0_Presence_1~Feeder_number, FD_Data)

lmfit_FD #Let's view what we just did

@@

tidy(lmfit_FD) #We can use the tidy() function to produce the output/ summary statistics

@@

#Checking model performance: FEEDER
check_model(lmfit_FD$fit)
```

```{r}
#More fitting with some other indicators
lmog_FD2<-linear_reg() %>% 
  set_engine("lm")

lmfit_FD2<- lmog_FD2 %>% #Training; outcome: Salmonella Positive and Main Predictor Feeder Number and Type (e.g., wood v. plastic)
  fit(Absence_0_Presence_1~Feeder_number+Feeder_Type, FD_Data)

lmfit_FD2 #Let's view what we just did

@@

tidy(lmfit_FD2) #We can use the tidy() function to produce the output/ summary statistics

@@

check_model(lmfit_FD2$fit)
```

```{r}
#Trying ALL predictors of Feeder dataset 
lmog_FD<- linear_reg() %>%
  set_engine("lm")

@@

tidy(lmall_FD) #We can use the tidy() function to produce the output/ summary statistics

plot_summs(lmall_FD, robust=TRUE)

check_model(lmall_FD$fit)
```

# 3. **Data Fitting/Statistical Analysis: Generalized Linear Model \[Picnic Table Trial\]**

Unlike linear regressions, generalized linear models (GLMs) are often more flexible and allow for change in inputs. Below I will run GLMs for both datasets and check each model.

```{r}
lmog<-linear_reg() %>% 
  set_engine("glm")

@@

tidy(lmfit_PT_glm)

@@

check_model(lmfit_PT_glm$fit)

@@

lmfit_PTglm1<- lmog_glm1 %>% #Training; outcome: Salmonella Positive and MPs: Pile, Table, and Date
  fit(Salmonella_Positive~Original_Persistence_Pile+Table+Date, PT_Data)

@@

tidy(lmfit_PTglm1)

@@

check_model(lmfit_PTglm1$fit)

@@

lmfit_PTglm2<- lmog_glm2 %>% #Training; outcome: Salmonella Pos and Main Predictor Table
  fit(Salmonella_Positive~Quadrant+Table, PT_Data)

check_model(lmfit_PTglm2$fit)
```

@@

tidy(lmfit_PTglm2)

@@

check_model(lmfit_PTglm2$fit)

@@

## a. **Generalized Linear Model: Picnic Table (Persistence & Prevalence)**
Again, this is an alternative way to run a glm.

```{r}
PT_GLM_PER <- glm(Salmonella_Positive ~ Date, data = PT_Data, family = binomial) #Persistence

@@

tidy(PT_GLM_PER)

PT_GLM_PREV <- glm(Salmonella_Positive ~ Sample_Type + as.factor(Date) + Table, data= PT_Data, family = binomial) #Prevalence

PT_GLM_PREV

tidy(PT_GLM_PREV)

lmall_PTGLM
```

# 4. **Data Fitting/Statistical Analysis: Generalized Linear Model [Feeder Trial]**

```{r}
lmog_FD<-linear_reg() %>% 
  set_engine("glm")

@@

@@

check_model(lmfit_FD_glm$fit)

@@

tidy(lmfit_FD_GLM)

@@

check_model(lmfit_FD_GLM$fit)

@@

tidy(lmall_FDGLM_all1)

check_model(lmall_FDGLM_all1$fit)
```

# **Generalized Linear Model: Feeder Data (Persistence)**

```{r}
FD_GLM_PER <- glm(Absence_0_Presence_1 ~ Collection_Date, data = FD_Data, family = binomial) #Persistence

@@

FD_GLM_PREV <- glm(Absence_0_Presence_1 ~ Feeder_Type + as.factor(Collection_Date), data= FD_Data, family = binomial) #Prevalence

FD_GLM_PREV
```

# **Time to event**

```{r}
library(finalfit)
library(forcats)
library(survival)

@@

# **Cox regression: Picnic Table Data**

```{r}
PT_Data$Date<- as.numeric(PT_Data$Date)

@@

coxph(Surv(Date, Salmonella_Positive) ~ Sample_Type, PT_Data) %>%
  tbl_regression(exp=TRUE)
```

\#**Cox regression: Feeder Data**

```{r}
library(lubridate)

### 2023-05-05 5. Manuscript/Manuscript.qmd
More updates

After collection, 100 uL of the sample vortexed in dulcitol selentite
was inoculated into 10 mL of Rappaport Vassiliadis (RV) broth. The broth
was incubated for 24 hours at 37c. The RV broth was then streaked onto
XLT-4 agar plates and incubated at 37c for 24 hours, then an additional
24 hours on the lab bench at room temperature. If colonies
morphometrically consistent with *Salmonella* were present, one colony
per plate was chosen and a stab in Luria Broth (LB) agar was created. To
confirm the presence of *Salmonella*, patch plates were made on
ChromAgar using the stabbed colonies. If the patches turned magenta,
that indicated positive *Salmonella* samples. A blue color indicated
something other than *Salmonella* [@Maurer2015].

@@

**Picnic Table Data Exploration and Visualization**

![Positive Salmonella Samples by Type and
Date](images/Picnic_Table_PosSamp-05.png)***Figure 1.*** Positive
*Salmonella* Samples by Sample Type and Date

![***Figure 2.*** Positive *Salmonella* Samples by Quadrant and
Date](images/Picnic_Table_Quad-02.png)

**Bird Feeder Data Exploration and Visualization**

![***Figure 3.*** Positive *Salmonella* Samples by feeder type and
date.](images/Feeder_PA.png)

## **Ibis Trial**

>>>>

After collection, dulcitol selenite was incubated overnight at 37°c,
before 100 uL of the sample vortexed in dulcitol selentite was
inoculated into 10 mL of Rappaport Vassiliadis (RV) broth. The broth was
incubated for 24 hours at 37°c. The RV broth was then streaked onto
XLT-4 agar plates and incubated at 37°c for 24 hours, then an additional
24 hours on the lab bench at room temperature. If colonies
morphometrically consistent with *Salmonella* were present, one colony
per plate was chosen and a stab in Luria Broth (LB) agar was created. To
confirm the presence of *Salmonella*, patch plates were made on
ChromAgar using the stabbed colonies. If the patches turned magenta,
that indicated positive *Salmonella* samples. A blue color indicated
something other than *Salmonella* [@Maurer2015].

@@

**Picnic Table Data Exploration and Visualization**
```{r, echo=FALSE, message= FALSE, fig.width=10, fig.height=8, fig.cap= "Figure. 1 Positive Salmonella Samples by Sample Type and Date"}

library(here)

knitr::include_graphics(here("4. Results/Picnic_Table_PosSamp.png"))
```

```{r, echo=FALSE, message= FALSE, fig.width=10, fig.height=8, fig.cap= "Figure. 2 Positive Salmonella Samples by Quadrant and Date"}

library(here)

knitr::include_graphics(here("4. Results/Picnic_Table_Quad.png"))
```

**Bird Feeder Data Exploration and Visualization**
```{r, echo=FALSE, message= FALSE, fig.width=10, fig.height=8, fig.cap= "Figure. 3 Positive Salmonella Samples by Feeder Type and Date"}

library(here)

knitr::include_graphics(here("4. Results/Feeder_PA.png"))
```

## **Picnic Table Trial**
